\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-09-10 manuscript.tex
edits
sas code included
edits through results

\textbf{Genotyping analysis---}To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations ($n = 32$) in which eight or more chromosomes were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between \emph{Hopscotch} frequency and environmental variables using the software BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (\emph{i.e.,} Bayes Factor; \citealt{Coop2010}). We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana}) genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP. SNP data from \citep{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA). Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  These data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}.

@@

\textbf{Phenotyping of \emph{parviglumis}---}To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte we conducted a phenotyping trial in which we germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} insertion is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites within the population where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with Captan fungicide (Southern Agricultural Insecticides Inc., Palmetto, Florida, USA) and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were planted into one-gallon pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer, which was supplemented with hand watering on extremely hot and dry days.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to those reported genome-wide in previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is less negative in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

@@

\textbf{Phenotyping---}To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion based on our initial genotyping. 
We measured tiller number and tillering index, the ratio of the sum of tiller lengths to plant height, for 206 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. 
We also measured culm diameter, a phenotype that differs between maize and teosinte \citep{Briggs2007} but is not thought to be affected by the \emph{Hopscotch} insertion. 
Phenotypic data are available at \url{http://dx.doi.org/10.6084/m9.figshare.776926}. 
Our plantings produced 82 homozygotes for the \emph{Hopscotch} insertion at \emph{tb1}, 104 heterozygotes, and 20 homozygotes lacking the insertion; these numbers do not deviate from expectations of Hardy-Weinberg equilibrium. 
After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find no significant correlation between genotype at the \emph{Hopscotch} insertion and tillering index (Fig. \ref{Fig4Boxplots}), tiller number, or culm diameter. 
Only on day 40 did we observe a weak but statistically insignificant ($p=0.0848$, $r^2$ = 0.02) correlation between tillering index and the \emph{Hopscotch} genotype, although in the opposite direction of that expected, with homozygotes for the insertion showing a higher tillering index.

>>>>

\textbf{Genotyping analysis---}To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations ($n = 32$) in which eight or more chromosomes were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between \emph{Hopscotch} frequency and environmental variables using the software BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (\emph{i.e.,} Bayes Factor; \citealt{Coop2010}). We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana}) genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional biallelic marker. SNP data from \citet{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA). Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  Environmental data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}.

@@

\textbf{Phenotyping of \emph{parviglumis}---}To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte we conducted a phenotyping trial in which we germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo; \citealt{Hufford2010}) where the \emph{Hopscotch} insertion is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites within the population where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with Captan fungicide (Southern Agricultural Insecticides Inc., Palmetto, Florida, USA) and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were planted into one-gallon pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer, which was supplemented with hand watering on extremely hot and dry days.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency within populations in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to genome-wide estimates from previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is closer to zero in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

@@

\textbf{Phenotyping---}To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion based on our initial genotyping. 
We measured tiller number and tillering index, the ratio of the sum of tiller lengths to plant height, for 206 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. 
We also measured culm diameter, a phenotype that differs between maize and teosinte \citep{Briggs2007} but is not thought to be affected by the \emph{Hopscotch} insertion. 
Phenotypic data are available at \url{http://dx.doi.org/10.6084/m9.figshare.776926}. 
Our plantings produced 82 homozygotes for the \emph{Hopscotch} insertion at \emph{tb1}, 104 heterozygotes, and 20 homozygotes lacking the insertion; these numbers do not deviate from expectations of Hardy-Weinberg equilibrium. 
After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find no significant correlation between genotype at the \emph{Hopscotch} insertion and tillering index (Fig. \ref{Fig4Boxplots}), tiller number, or culm diameter. 
Only on day 40 did we observe a weak but statistically insignificant ($r^2= 0.02$, $p=0.0848$) correlation between tillering index and the \emph{Hopscotch} genotype, although in the opposite direction of that expected, with homozygotes for the insertion showing a higher tillering index.
